An open and continuously updated fern tree of life (FTOL)

### 2022-02-13 ms/manuscript.Rmd
Use "major_clade" instead of "plot_group"
Update text about divergence times

# Load functions
source(here::here("R/ms_functions.R"))

@@

# Helper function to add plot_group:
# a "rank" that may be at order or suborder level for plotting
add_plot_group <- function(data) {
	data %>%
		mutate(
			plot_group = coalesce(suborder, order),
			plot_group = case_when(
				plot_group %in% c("Ophioglossales", "Psilotales") ~ 
					"Ophioglossales + Psilotales",
				TRUE ~ plot_group
			)
		)
}

# Tally FTOL sampling by rank
ftol_sampling_by_rank <-
	sanger_sampling %>%
	filter(outgroup == FALSE) %>%
	select(-outgroup) %>%
	# add plot_group
	add_plot_group() %>%
	# drop suborder
	select(-suborder) %>%
	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
	group_by(rank) %>%
	count(name) %>%
	ungroup

# Make list of all accepted species
accepted_species <-
	pteridocat %>% 
	# Only keep accepted names
	filter(str_detect(taxonomicStatus, "accepted")) %>%
	# Only keep species
	# FIXME: need to add rank for newly added names in pteridocat_maker
	# then don't need is.na()
	filter(taxonRank == "species" | is.na(taxonRank)) %>%
	select(scientificName) %>%
	mutate(
		rgnparser::gn_parse_tidy(scientificName) %>% 
			select(species = canonicalsimple),
		species = str_replace_all(species, " ", "_")
	) %>%
	# Map on higher level taxonomy
	mutate(genus = str_split(scientificName, " ") %>% map_chr(1)) %>%
	left_join(
		select(ppgi_taxonomy, genus, family, 
					 suborder, order, class), by = "genus"
	) %>%
	# FIXME: update class for nothogenera in PPG
	mutate(class = case_when(
		str_detect(
			scientificName, "Schizoloma|Phlebosia|Cystocarpium") ~ "Polypodiopsida",
		TRUE ~ class
	)) %>%
	# Drop lycophytes
	assert(not_na, class) %>%
	filter(class == "Polypodiopsida") %>%
	select(-class) %>%
	# Add plot_group
	add_plot_group() %>%
	# drop suborder
	select(-suborder)

@@

# Format for text in MS
tax_coverage <-
	coverage_data %>%
	filter(rank == "plot_group") %>%
	arrange(coverage) %>%
	transmute(clade = name, coverage = percent(coverage, accuracy = 0.1)) %>%
	split(.$clade)

@@

We estimate the crown age of ferns to be `r fern_age` ma.
Ages of other major crown groups are leptosporangiates (`r lepto_age` ma), Polypodiales (`r polypodiales_age` ma), eupolypods I (`r eupoly_i_age` ma), and eupolypods II (`r eupoly_ii_age` ma).
Except for eupolypods II, these ages are all slightly older than those reported in the most recent global fern phylogeny [@Testo2016a].
Ages of fern families differed from previous studies, with some families estimated as having older crown ages while some had younger (`r figure("divtimes")`).

@@

plastome_gleich_genera_n <-
	bb_sampling %>%
	filter(plot_group == "Gleicheniales") %>%
	pull(genus) %>%
	n_distinct()
```

@@

Although we recovered slightly older ages for ferns and other large clades relative to the most recent globally sampled fern phylogeny [@Testo2016a], our estimated ages for families do not show a clear pattern, with some older and some younger than other studies (`r figure("divtimes")`).
This may be due to the fact that we used a completely revised set of fossil calibration points relative to previous studies.
Several of the recent studies analyzing divergence times for global fern phylogenies all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
However, we found several of these calibration points were not appropriately applied due to differences in taxonomic concepts between the original fossil publication and the phylogeny in which they were used.
Moreover, our set of calibration points more than doubles the number used by previous studies.
Therefore, it is perhaps not surprising that we obtained different estimates for the age of many families, whereas previous studies tended to estimate similar ages by re-using the same (or very similar), more limited set of calibration points (SI Fig 1).

@@

# Make tibble of backbone tips with higher-level taxonomy
backbone_taxonomy <-
	tibble(species = get_tips_in_ape_plot_order(plastome_tree_rooted)) %>%
	# Add higher taxonomy
	left_join(sanger_sampling, by = "species") %>%
	# Add column for phylogenetic order
	mutate(phy_order = seq_len(nrow(.))) %>%
	# Outgroup lacks family data, so call "og"
	mutate(
		family = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ family),
		plot_group = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ plot_group)
	)
```

@@

# Make exemplar set of species in backbone representing plot groups
plot_group_exemp_taxa <-
	backbone_taxonomy %>%
	# Slice to one exemplar species per major clade
	group_by(plot_group) %>%
	slice(1) %>%
	ungroup() %>%
	# Slicing arranges alphabetically, so put back in (unladderized) phylo order
	arrange(phy_order) %>%
	# Drop outgroup
	filter(family != "og")

@@

clade_cols <-
	Polychrome::green.armytage.colors() %>%
	extract(green_army_use_cols) %>%
	extract(seq_len(nrow(plot_group_exemp_taxa))) %>%
	set_names(plot_group_exemp_taxa$plot_group)

# Format coverage at order-ish level for plotting
b <-
	coverage_data %>%
	filter(rank == "plot_group") %>%
	left_join(plot_group_exemp_taxa, by = c(name = "plot_group")) %>%
	arrange(phy_order) %>%
	rename(clade = name) %>%
	mutate(
		clade = fct_reorder(clade, phy_order) %>%
			fct_rev(),
		coverage_text = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")
	) %>%
	ggplot(aes(x = coverage, y = clade, fill = clade)) +
	geom_col(color = "grey10", size = 0.2) +
	geom_text(
		aes(label = coverage_text),
		hjust = 0,
		nudge_x = 0.01,
		size = 6*.36 # 6 pt font
	) +
	scale_x_continuous(
		labels = percent,
		limits = c(0,1),
		expand = expansion(mult = c(0, 0))
	) +
	scale_fill_manual(values = clade_cols) +
	labs(x = "Coverage") +
	theme_bw() +
	theme(
		axis.text.y = element_text(size = 6),
		axis.title.y = element_blank(),
		legend.position = "none",
		plot.background = element_rect(fill = "transparent", colour = NA),
		panel.grid.major.y = element_blank(),
		panel.grid.minor.x = element_blank()
	)

@@

# Make tibble for annotating major clades
clade_tibble <-
	tibble(
		Polypodiineae = c("Leucostegia_truncata", "Didymochlaena_fijiensis"),
		Aspleniineae = c("Acystopteris_taiwaniana", "Rhachidosorus_pulcher"),
		Dennstaedtiineae = c("Monachosorum_flagellare", "Dennstaedtia_auriculata"),
		Pteridineae = c("Coniogramme_affinis", "Pteris_dispar"),
		Lindsaeineae = c("Cystodium_sorbifolium", "Odontosoria_retusa"),
		Saccolomatineae = c("Saccoloma_inaequale", "Orthiopteris_campylura"),
		Cyatheales = c("Cyathea_bipinnatifida", "Culcita_macrocarpa"),
		Salviniales = c("Marsilea_nashii", "Salvinia_oblongifolia"),
		Schizaeales = c("Schizaea_elegans", "Lygodium_palmatum"),
		Gleicheniales = c("Matonia_foxworthyi", "Sticherus_bifidus"),
		Hymenophyllales = c("Cephalomanes_javanicum", "Hymenophyllum_tortuosum"),
		Marattiales = c("Angiopteris_evecta", "Danaea_erecta"),
		`Ophioglossales + Psilotales` = c("Psilotum_nudum", "Sceptridium_nipponicum"),
		Osmundales = c("Osmundastrum_cinnamomeum", "Todea_barbara"),
		Equisetales = c("Equisetum_bogotense", "Equisetum_hyemale")
	) %>%
	pivot_longer(names_to = "plot_group", values_to = "taxa", cols = everything()) %>%
	arrange(plot_group) %>%
	group_by(plot_group) %>%
	summarize(taxa = list(taxa)) %>%
	mutate(
		mrca = map_chr(taxa, ~getMRCA(phy, .))
	) %>%
	transmute(
		plot_group,
		mrca,
		dummy = ""
	) %>%
	mutate(mrca = parse_number(mrca))

a <- ggtree(
	phy, 
	layout = "circular", right = FALSE, size = 0.1,
) +
	geom_cladelab(
		data = clade_tibble,
		mapping = aes(
			node = mrca,
			color = plot_group,
			label = dummy
		),
		offset = 0.1,
		barsize = 2) +
	scale_color_manual(values = clade_cols) +
	theme(legend.position = "none")

@@

suborder_labels_singletons <-
	family_exemp %>% 
	filter(outgroup == FALSE) %>%
	group_by(plot_group) %>%
	select(plot_group, family) %>% 
	filter(n_distinct(family) == 1) %>%
	ungroup() %>%
	left_join(phy_tip, by = c(family = "label")) %>%
	select(plot_group, mrca = node)

suborder_labels <-
	family_exemp %>% 
	filter(outgroup == FALSE) %>%
	group_by(plot_group) %>%
	select(plot_group, family) %>% 
	filter(n_distinct(family) > 1) %>%
	summarize(taxa = list(family)) %>%
	mutate(
		mrca = map_chr(taxa, ~getMRCA(phy, .))
	) %>%
	transmute(
		plot_group,
		mrca
	) %>%
	mutate(mrca = parse_number(mrca)) %>%
	bind_rows(suborder_labels_singletons) %>%
	mutate(
		clade_label = case_when(
			plot_group == "Aspleniineae" ~ "Aspleniineae\n(eupolypods II)",
			plot_group == "Polypodiineae" ~ "Polypodiineae\n(eupolypods I)",
			TRUE ~ plot_group
		)
	)

@@

set.seed(123)
# Use the ggtree::`%<+%` operator to map the 
# bootstrap values onto the ggtree object
backbone_plot <- 
	ggtree(phy) %<+% bs_tibble +
	# now we can show BS values using geom_text()
	geom_text(aes(label = bootstrap), hjust = -.25, size = 2.5) +
	# add tip labels
	geom_tiplab(aes(label = label)) +
	# highlight major clades
	geom_hilight(
		data = suborder_labels,
		mapping = aes(
			node = mrca,
			fill = plot_group
		),
		alpha = 0.2,
		align = "right",
		extend = 30,
		gradient.direction = "rt",
		roundrect = TRUE,
		gradient.length.out = 2) +
	scale_fill_manual(values = clade_cols) +
	# label major clades
	geom_cladelab(
		data = suborder_labels,
		mapping = aes(
			node = mrca,
			label = clade_label
		),
		offset = 28,
		align = TRUE,
		extend = 0.3,
		hjust = 0,
		barcolour = "grey20",
		barsize = 2) +
	# Label other groups
	geom_text_repel(
		data = other_clade_labels %>%
			filter(clade_label == "Polypodiidae\n(leptosporangiates)"),
		aes(x = x, y = y, label = clade_label),
		hjust = 0,
		xlim = c(NA, 10),
		ylim = c(15, NA),
		color = "grey20",
		segment.color = "grey20"
	) +
	geom_text_repel(
		data = other_clade_labels %>%
			filter(clade_label == "eupolypods"),
		aes(x = x, y = y, label = clade_label),
		hjust = 0,
		xlim = c(NA, 64),
		ylim = c(32, NA),
		color = "grey20",
		segment.color = "grey20"
	) +
	geom_text_repel(
		data = other_clade_labels %>%
			filter(clade_label == "Polypodiales"),
		aes(x = x, y = y, label = clade_label),
		hjust = 0,
		xlim = c(NA, 48),
		ylim = c(24, NA),
		color = "grey20",
		segment.color = "grey20"
	) +
	xlim(0, 180) +
	coord_cartesian(clip = "off") +
	theme(legend.position = "none")
```

@@

**`r figure("divtimes")`**. Crown age of fern families (mya) estimated by selected studies.
For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a].
For other studies, point indicates best (most likely) estimate [@Schuettpelz2007b\; this study].

>>>>

# Load functions
source(here::here("R/ms_functions.R"))
source(here::here("R/functions.R"))

@@

# Helper function to add major_clade:
# a "rank" that may be at order or suborder level for plotting
add_major_clade <- function(data) {
	data %>%
		mutate(
			major_clade = coalesce(suborder, order),
			major_clade = case_when(
				major_clade %in% c("Ophioglossales", "Psilotales") ~ 
					"Ophioglossales + Psilotales",
				TRUE ~ major_clade
			)
		)
}

# Tally FTOL sampling by rank
ftol_sampling_by_rank <-
	sanger_sampling %>%
	filter(outgroup == FALSE) %>%
	select(-outgroup) %>%
	# add major_clade
	add_major_clade() %>%
	# drop suborder
	select(-suborder) %>%
	pivot_longer(names_to = "rank", values_to = "name", -species) %>%
	group_by(rank) %>%
	count(name) %>%
	ungroup

# Make list of all accepted species
accepted_species <-
	pteridocat %>% 
	# Only keep accepted names
	filter(str_detect(taxonomicStatus, "accepted")) %>%
	# Only keep species
	# FIXME: need to add rank for newly added names in pteridocat_maker
	# then don't need is.na()
	filter(taxonRank == "species" | is.na(taxonRank)) %>%
	select(scientificName) %>%
	mutate(
		rgnparser::gn_parse_tidy(scientificName) %>% 
			select(species = canonicalsimple),
		species = str_replace_all(species, " ", "_")
	) %>%
	# Map on higher level taxonomy
	mutate(genus = str_split(scientificName, " ") %>% map_chr(1)) %>%
	left_join(
		select(ppgi_taxonomy, genus, family, 
					 suborder, order, class), by = "genus"
	) %>%
	# FIXME: update class for nothogenera in PPG
	mutate(class = case_when(
		str_detect(
			scientificName, "Schizoloma|Phlebosia|Cystocarpium") ~ "Polypodiopsida",
		TRUE ~ class
	)) %>%
	# Drop lycophytes
	assert(not_na, class) %>%
	filter(class == "Polypodiopsida") %>%
	select(-class) %>%
	# Add major_clade
	add_major_clade() %>%
	# drop suborder
	select(-suborder)

@@

# Format for text in MS
tax_coverage <-
	coverage_data %>%
	filter(rank == "major_clade") %>%
	arrange(coverage) %>%
	transmute(clade = name, coverage = percent(coverage, accuracy = 0.1)) %>%
	split(.$clade)

@@

We estimate the crown age of ferns to be `r fern_age` ma.
Ages of other major crown groups are leptosporangiates (`r lepto_age` ma), Polypodiales (`r polypodiales_age` ma), eupolypods I (`r eupoly_i_age` ma), and eupolypods II (`r eupoly_ii_age` ma).
Except for eupolypods II, these ages are all slightly older than those reported in the most recent global fern phylogeny [@Testo2016a].
Estimated stem ages of fern families were generally older than previous studies. (`r figure("divtimes")`).

@@

plastome_gleich_genera_n <-
	bb_sampling %>%
	filter(major_clade == "Gleicheniales") %>%
	pull(genus) %>%
	n_distinct()
```

@@

We recovered older crown ages for ferns and other large clades as well as older stem ages for families relative to previous studies (`r figure("divtimes")`).
This may be due to the fact that we used a completely revised set of fossil calibration points relative to previous studies.
Several of the recent studies analyzing divergence times for global fern phylogenies all used a similar set of 24--26 fossil calibration points [@Schuettpelz2007b; @Testo2016a] or secondary calibration points based on studies using this set [@Rothfels2015a].
However, we found several of these calibration points were not appropriately applied due to differences in taxonomic concepts between the original fossil publication and the phylogeny in which they were used.
Moreover, our set of calibration points more than doubles the number used by previous studies.
It is likely that this expanded sampling has yielded older estimates for the stem age of many families, whereas previous studies tended to estimate similar ages by re-using the same (or very similar), more limited set of calibration points (SI Fig 1).

@@

# Make tibble of backbone tips with higher-level taxonomy
backbone_taxonomy <-
	tibble(species = get_tips_in_ape_plot_order(plastome_tree_rooted)) %>%
	# Add higher taxonomy
	left_join(sanger_sampling, by = "species") %>%
	# Add column for phylogenetic order
	mutate(phy_order = seq_len(nrow(.))) %>%
	# Outgroup lacks family data, so call "og"
	mutate(
		family = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ family),
		major_clade = case_when(
			outgroup == TRUE ~ "og",
			TRUE ~ major_clade)
	)
```

@@

# Make exemplar set of species in backbone representing plot groups
major_clade_exemp_taxa <-
	backbone_taxonomy %>%
	# Slice to one exemplar species per major clade
	group_by(major_clade) %>%
	slice(1) %>%
	ungroup() %>%
	# Slicing arranges alphabetically, so put back in (unladderized) phylo order
	arrange(phy_order) %>%
	# Drop outgroup
	filter(family != "og")

@@

clade_cols <-
	Polychrome::green.armytage.colors() %>%
	extract(green_army_use_cols) %>%
	extract(seq_len(nrow(major_clade_exemp_taxa))) %>%
	set_names(major_clade_exemp_taxa$major_clade)

# Format coverage at order-ish level for plotting
b <-
	coverage_data %>%
	filter(rank == "major_clade") %>%
	left_join(major_clade_exemp_taxa, by = c(name = "major_clade")) %>%
	arrange(phy_order) %>%
	rename(clade = name) %>%
	mutate(
		clade = fct_reorder(clade, phy_order) %>%
			fct_rev(),
		coverage_text = glue("{number(n, accuracy = 1)}/{number(n_accepted, accuracy = 1)}")
	) %>%
	ggplot(aes(x = coverage, y = clade, fill = clade)) +
	geom_col(color = "grey10", size = 0.2) +
	geom_text(
		aes(label = coverage_text),
		hjust = 0,
		nudge_x = 0.01,
		size = 6*.36 # 6 pt font
	) +
	scale_x_continuous(
		labels = percent,
		limits = c(0,1),
		expand = expansion(mult = c(0, 0))
	) +
	scale_fill_manual(values = clade_cols) +
	labs(x = "Coverage") +
	theme_bw() +
	theme(
		axis.text.y = element_text(size = 6),
		axis.title.y = element_blank(),
		legend.position = "none",
		plot.background = element_rect(fill = "transparent", colour = NA),
		panel.grid.major.y = element_blank(),
		panel.grid.minor.x = element_blank()
	)

@@

# Make tibble for annotating major clades
clade_tibble <-
	tibble(
		Polypodiineae = c("Leucostegia_truncata", "Didymochlaena_fijiensis"),
		Aspleniineae = c("Acystopteris_taiwaniana", "Rhachidosorus_pulcher"),
		Dennstaedtiineae = c("Monachosorum_flagellare", "Dennstaedtia_auriculata"),
		Pteridineae = c("Coniogramme_affinis", "Pteris_dispar"),
		Lindsaeineae = c("Cystodium_sorbifolium", "Odontosoria_retusa"),
		Saccolomatineae = c("Saccoloma_inaequale", "Orthiopteris_campylura"),
		Cyatheales = c("Cyathea_bipinnatifida", "Culcita_macrocarpa"),
		Salviniales = c("Marsilea_nashii", "Salvinia_oblongifolia"),
		Schizaeales = c("Schizaea_elegans", "Lygodium_palmatum"),
		Gleicheniales = c("Matonia_foxworthyi", "Sticherus_bifidus"),
		Hymenophyllales = c("Cephalomanes_javanicum", "Hymenophyllum_tortuosum"),
		Marattiales = c("Angiopteris_evecta", "Danaea_erecta"),
		`Ophioglossales + Psilotales` = c("Psilotum_nudum", "Sceptridium_nipponicum"),
		Osmundales = c("Osmundastrum_cinnamomeum", "Todea_barbara"),
		Equisetales = c("Equisetum_bogotense", "Equisetum_hyemale")
	) %>%
	pivot_longer(names_to = "major_clade", values_to = "taxa", cols = everything()) %>%
	arrange(major_clade) %>%
	group_by(major_clade) %>%
	summarize(taxa = list(taxa)) %>%
	mutate(
		mrca = map_chr(taxa, ~getMRCA(phy, .))
	) %>%
	transmute(
		major_clade,
		mrca,
		dummy = ""
	) %>%
	mutate(mrca = parse_number(mrca))

a <- ggtree(
	phy, 
	layout = "circular", right = FALSE, size = 0.1,
) +
	geom_cladelab(
		data = clade_tibble,
		mapping = aes(
			node = mrca,
			color = major_clade,
			label = dummy
		),
		offset = 0.1,
		barsize = 2) +
	scale_color_manual(values = clade_cols) +
	theme(legend.position = "none")

@@

suborder_labels_singletons <-
	family_exemp %>% 
	filter(outgroup == FALSE) %>%
	group_by(major_clade) %>%
	select(major_clade, family) %>% 
	filter(n_distinct(family) == 1) %>%
	ungroup() %>%
	left_join(phy_tip, by = c(family = "label")) %>%
	select(major_clade, mrca = node)

suborder_labels <-
	family_exemp %>% 
	filter(outgroup == FALSE) %>%
	group_by(major_clade) %>%
	select(major_clade, family) %>% 
	filter(n_distinct(family) > 1) %>%
	summarize(taxa = list(family)) %>%
	mutate(
		mrca = map_chr(taxa, ~getMRCA(phy, .))
	) %>%
	transmute(
		major_clade,
		mrca
	) %>%
	mutate(mrca = parse_number(mrca)) %>%
	bind_rows(suborder_labels_singletons) %>%
	mutate(
		clade_label = case_when(
			major_clade == "Aspleniineae" ~ "Aspleniineae\n(eupolypods II)",
			major_clade == "Polypodiineae" ~ "Polypodiineae\n(eupolypods I)",
			TRUE ~ major_clade
		)
	)

@@

set.seed(123)
# Use the ggtree::`%<+%` operator to map the 
# bootstrap values onto the ggtree object
backbone_plot <- 
	ggtree(phy) %<+% bs_tibble +
	# now we can show BS values using geom_text()
	geom_text(aes(label = bootstrap), hjust = -.25, size = 2.5) +
	# add tip labels
	geom_tiplab(aes(label = label)) +
	# highlight major clades
	geom_hilight(
		data = suborder_labels,
		mapping = aes(
			node = mrca,
			fill = major_clade
		),
		alpha = 0.2,
		align = "right",
		extend = 30,
		gradient.direction = "rt",
		roundrect = TRUE,
		gradient.length.out = 2) +
	scale_fill_manual(values = clade_cols) +
	# label major clades
	geom_cladelab(
		data = suborder_labels,
		mapping = aes(
			node = mrca,
			label = clade_label
		),
		offset = 28,
		align = TRUE,
		extend = 0.3,
		hjust = 0,
		barcolour = "grey20",
		barsize = 2) +
	# Label other groups
	geom_text_repel(
		data = other_clade_labels %>%
			filter(clade_label == "Polypodiidae\n(leptosporangiates)"),
		aes(x = x, y = y, label = clade_label),
		hjust = 0,
		xlim = c(NA, 10),
		ylim = c(15, NA),
		color = "grey20",
		segment.color = "grey20"
	) +
	geom_text_repel(
		data = other_clade_labels %>%
			filter(clade_label == "eupolypods"),
		aes(x = x, y = y, label = clade_label),
		hjust = 0,
		xlim = c(NA, 64),
		ylim = c(32, NA),
		color = "grey20",
		segment.color = "grey20"
	) +
	geom_text_repel(
		data = other_clade_labels %>%
			filter(clade_label == "Polypodiales"),
		aes(x = x, y = y, label = clade_label),
		hjust = 0,
		xlim = c(NA, 48),
		ylim = c(24, NA),
		color = "grey20",
		segment.color = "grey20"
	) +
	xlim(0, 180) +
	coord_cartesian(clip = "off") +
	theme(legend.position = "none")
```

@@

**`r figure("divtimes")`**. Stem age of fern families (mya) estimated by selected studies.
For studies that used methods with confidence intervals, error bars indicate lower and upper 95% highest posterior density levels and point indicates median [@Rothfels2015a; @Testo2016a].
For other studies, point indicates best (most likely) estimate [@Schuettpelz2007b\; this study].
